Emitter Source Geolocation from Imparted Rotor Blade Modulation

### 2016-10-11 Background.tex
update files

\chapter{Literature Review and Geometric Models}

@@

\section{Literature Review}
%this is the literature review section on what rbm is and current research in the area
The rotating blades of the helicopter modulate signals that hit the blades by reflecting them. The reflected signal, through the doppler effect, become shifted in frequency in relationship to the velocity of the reflection point (refs?). The rotor blade modulation (RBM) has a complex form and is influenced by several object parameters. These parameters range from the positioning of a signal source to the physical dimensions of the blades. The resulting modulated signal has been used in radar applications relating to the classification of any aircraft with a rotor blade, propeller or even jet engines (refs?). The doppler shifted signal has also been shown to have a parasitic effect on GPS anti-jamming algorithms (refs?). The role of RBM is based on the specific doppler shift associated with where a dominant reflection occurs along the length of the of the blade, and the associated elevation angle of the transmitter. The modulated signal will then be used to create a mapping between the amount of doppler present and the distance of a transmitter on the ground.

>>>>

\chapter{Literature Review and Geometric Models} \label{ch:background}

@@

\section{Literature Review}
%this is the literature review section on what rbm is and current research in the area
The rotating blades of the helicopter modulate signals that hit the blades by reflecting them. The reflected signal, through the doppler effect, become shifted in frequency in relationship to the velocity of the reflection point (Yaman Zhang 2000). The rotor blade modulation (RBM) has a complex form and is influenced by several object parameters. These parameters range from the positioning of a signal source to the physical dimensions of the blades.

The resulting modulated signal has been used in RADAR applications. First to determine the specific characteristics of a noncooperative helicopter using time frequency analysis in (Barry D. Bullard 1991) that defines a specific signature for each parameter. The ability to detect a hovering helicopter based on the blade flashes caused by reflected RADAR pulses off the blades is defined in (J Misiurewicz 1997). The micro-doppler produced by the RADAR reflection is used in (Thayaparan, T. 2007) to determine the rotation rate of the main and tail rotor blades using the wavelet transform.

The doppler shifted signal effects the performance of common adaptive GPS anti-jamming algorithms located on a helicopter platform, unless the covariance estimation is completed within a few milliseconds. In iterative approaches the non-stationary signal characteristics cause degradation as well (F. Barbiero 2009). The affect of RBM on the BER of communications is examined in (Yaman Zhang 2000). Where effects of multi path and doppler cause large bit error rates at small elevation angles and large scattering coefficients described in their model.

The role of RBM is based on the specific doppler shift associated with where a dominant reflection occurs along the length of the of the blade, and the associated elevation angle of the transmitter. The modulated signal will then be used to create a mapping between the amount of doppler present and the distance of a transmitter on the ground.

### 2016-10-11 Conclusion.tex
update files

\chapter{Conclusion and Future Work}

@@

\section{Future Improvements}

>>>>

\chapter{Conclusion and Future Work} \label{ch:conclusion}

@@

Over the course of this thesis a three dimensional ray tracer application was built in C++ to produce a RBM signal according to a variety of physical parameters. The physical parameters associated with a helicopter blade and the positioning of transmitter and receiver hardware were swept independently to build knowledge on how they affect the resulting signal. That data was then processed using time frequency analysis to produce a doppler envelope. An envelope of each parameter step was used to determine an overall trend. Then with those trends an algorithm was derived to find the azimuth and elevation angle of the transmitter with relationship to the helicopter platform. The algorithm accurately finds the azimuth angle and elevation using the data collected in the simulation environment with almost no error in the azimuth estimation and less than 2\% error in elevation angle when blade pitch in 0\textdegree.

\section{Future Work}
%Real world experiment needs to be conducted starting with small scale and moving towards a large scale implementation on a helicopter platform.
The first thing that needs to be done in continuing this research is to perform a real world experiment to officially evaluate the ray tracing application. To do this appropriate hardware needs to be defined in the form of a software defined radio for both the receiver and transmitter along with a software workflow such as GNU Radio. Next a small scale environment would need to be set up, including the above hardware along with some form of rotating metal blade or thin plate rotating at a sufficient rate to produce a doppler spread similar to those defined in the simulation. After that data is collected and analyzed a full scale experiment would need to be conducted including a real helicopter platform and an environment with large expanses of flat ground.

%Evaluate the estimation techniques on the real world collected data to see if they are still applicable.
Once the real world experiments have been conducted, and are shown to be similar to the signal produced by the simulation, the estimation techniques defined in this thesis can be evaluated on the real signal to determine if they are sufficient in practice.

%Improve propagation model to better represent the real world conditions, material, more physical effects on RF, near field problems etc.
After the physical validation the ray tracing application can be modified to include more physical phenomenon. The other improvement would be to introduce more objects into the scene model which would require modification to the tracing algorithm to include ray casting distributions in their particular direction.

%Improve computation speed by employing multithreading or computation on GPU Architecture for reduced simulation time, not necessary but would be more convenient.
Another improvement on the simulation would be to increase the speed of computation. This could be done by implementing multi-threading or, since raytracing is heavily parallelizable, a GPU based architecture. The multi-threading option would require minor code modifications but realizing the simulation on a GPU would require a full rewrite of the software to take advantage of the GPU's abilities.

%Introduce a pitch parameter to the estimation equation based on empirical data, so that if the blades are pitched the doppler can be adjusted accordingly.
The current estimation does not take into account the physical pitch value when correcting for pitch and might not cover all the cases that pitch can effect. Therefore further investigation is needed into how the pitch affects the received signal for improved azimuth and elevation estimation.

%Because there is more information in the received envelopes besides the maximum and minimum the specific shape could be used to better estimate azimuth and elevation without the need to complete a full revolution as described in the proposed method.
The estimation methods described in this thesis rely only on the maximum and minimum values of the doppler envelopes calculated but their shape could also be of particular interest as well. When varying the parameters of azimuth and elevation the shape of the envelope changes in a distinctive manner, and by using a training algorithm the position of the transmitter might be able to be tracked in time by evaluating the change in shape. where currently if we want to locate the transmitter in a successive fashion a full revolution of the helicopter platform is needed to resolve position ambiguity.

### 2016-10-11 Introduction.tex
update files

\section{Contribution}
The main contributions of this thesis are as follows. A 3D Ray Tracing software is provided to simulate the RF propagation in the presence of a helicopter rotor blade. The software provides insight into the complex interaction between the rotor blade shape and the placement of the receiver and transmitters. The second contribution is the ability to use the imparted rotor blade modulation to estimate the azimuth and elevation angles with respect to the location of the rotating blades.

\section{Organization}
In Chapter 2 we present the background and theory behind rotor blade modulation. we then formulate the basic geometry behind the RBM effect.

In Chapter 3 we Introduce the radio propagation simulation in several sections. First is the RF propagation model which defines the physical effects that are simulated within the software. Next the objects of the scene model, as well as the dimensionality, are defined in software to be representative of their physical  dimensions and functionality. Lastly we describe the core ray tracing engine that traces the scene in successive frames to build the receiver output.

In Chapter 4 we present the result of the parameter swept simulations of the ray tracing software. Next the received signals from the simulation are analyzed using several DSP techniques to define the signal characteristics that will be used in the next chapter for estimation.

In Chapter 5 Those characteristics are then used to estimate the azimuth and elevation angle of the transmitter based on the given rotor and receiver parameters. Those estimates are then compared with the original position of the transmitter to determine the methods accuracy.

In Chapter 6 the summery of the results are detailed along with the current limitations. Lastly future improvements on the work will be discussed.

>>>>

\section{Contribution}
The main contributions of this thesis are as follows. A 3D Ray Tracing software is provided to simulate the RF propagation in the presence of a helicopter rotor blade. The software provides insight into the complex interaction between the rotor blade shape and the placement of the receiver and transmitters. The second contribution is an in depth analysis of how several different variables affect the modulation on the received signal. And lastly, techniques to use the imparted rotor blade modulation to estimate the azimuth and elevation angles with respect to the location of the rotating blades.

\section{Organization}
In Chapter \ref{ch:background} we present the background and theory behind rotor blade modulation. we then formulate the basic geometry behind the RBM effect.

In Chapter \ref{ch:radio_propagation} we Introduce the radio propagation simulation in several sections. First is the RF propagation model which defines the physical effects that are simulated within the software. Next the objects of the scene model, as well as the dimensionality, are defined in software to be representative of their physical  dimensions and functionality. Lastly we describe the core ray tracing engine that traces the scene in successive frames to build the receiver output.

In Chapter \ref{ch:simulations} we present the result of the parameter swept simulations of the ray tracing software. Next the received signals from the simulation are analyzed using several DSP techniques to define the signal characteristics that will be used in the next chapter for estimation.

In Chapter  \ref{ch:results} Those characteristics are then used to estimate the azimuth and elevation angle of the transmitter based on the given rotor and receiver parameters. Those estimates are then compared with the original position of the transmitter to determine the methods accuracy.

In Chapter \ref{ch:conclusion} the summery of the results are detailed along with the current limitations. Lastly future improvements on the work will be discussed.

### 2016-10-11 MSc. Thesis Template.tex
update files

% The Acknowledgements page, for thanking everyone
\acknowledgements{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics
I would like to thank my thesis advisor Dr. Tamal Bose for the opportunity to work in his lab, and for the guidance through graduate school. I would also like to thank Matt Kruse and Charlie Cooper for coming up the research topic and support through the process. I would also like to thank the Broadband Wireless Access \& Applications Center and Rincon Research Corporation for supporting my research.
}
\clearpage  
% End of the Acknowledgements

@@

}
%% ----------------------------------------------------------------
% End of the pre-able, contents and lists of things
% The Abstract Page
\addtotoc{Abstract}  % Add the "Abstract" page entry to the Contents
\abstract{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics
In RF communications with a rotorcraft, such as a helicopter, the rotor blades can impart a modulation onto the received signal called Rotor Blade Modulation (RBM). This modulation is caused by the reflection of a signal off the rotating blades. The reflected signal is Doppler shifted based on where the signal is reflected along the length of the blade as well as the angle between the axis of rotation and the emitter. RBM is known to degrade the performance of RF communications on rotorcraft and can be used in RADAR applications to detect and classify aircraft, but there is little on its usefulness in other areas.This project will look at the ability to utilize the RBM phenomenon on the rotorcraft itself to geo-locate and track a signal emitter on the ground. To do this a RF ray tracing program was developed in C++ to produce a simulation of the RBM.
}

@@

\include{Simulation_Data}

\include{Matlab_Code}

>>>>

% The Acknowledgements page, for thanking everyone
\acknowledgements{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics
I would like to thank my thesis advisor Dr. Tamal Bose for the opportunity to work in his lab, and for the guidance through graduate school. I would also like to thank Matt Kruse and Charlie Cooper for coming up the research topic and support through the process. I would also like to thank the Broadband Wireless Access \& Applications Center and Rincon Research Corporation for supporting this research.
}
\clearpage  
% End of the Acknowledgements

@@

}
%% ----------------------------------------------------------------
% End of the pre-able, contents and lists of things
% The Abstract Page
\addtotoc{Abstract}  % Add the "Abstract" page entry to the Contents
\abstract{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics
In RF communications with a rotorcraft, such as a helicopter, the rotor blades can impart a modulation onto the received signal called Rotor Blade Modulation (RBM). This modulation is caused by the reflection of a signal off the rotating blades. The reflected signal is Doppler shifted based on where the signal is reflected along the length of the blade as well as the angle between the axis of rotation and the emitter. RBM is known to degrade the performance of RF communications on rotorcraft and can be used in RADAR applications to detect and classify aircraft, but there is little on its usefulness in other areas. This thesis looks at the ability to utilize the RBM phenomenon on the rotorcraft itself to geo-locate and track a signal emitter on the ground.

To do this a 3D RF ray tracing program was developed in C++ to produce simulations of RBM signals. The developed program is based on optical ray tracing algorithms with modified physical propagation effects for RF signals, and swapping lights and cameras for RF transmitters and receivers respectively. The ray tracer was then run over a realistic set of physical parameters to determine their effects on the received signal, this includes transmitter azimuth and elevation angle, receiver position, blade pitch, etc. along with their combinations. From the simulations the azimuth and elevation angle, of the transmitter, produce predictable modulations on the received signal.  Based on the trends in the signal's modulation a DSP algorithm was distilled down that accurately determines the azimuth and elevation angle of the transmitter from simulated signal data.
}

@@

%\include{Simulation_Data}

%\include{Matlab_Code}

### 2016-10-11 Radio_Propagation_Simulation.tex
update files

\chapter{Radio Propagation Simulation}

\section{Introduction}
%introduce the section and what will be discussed here
The simulation software is broken down into three main components, the radio wave propagation model, the scene model, and the ray tracing engine. The wave propagation model is a collection of equations that describe the physical phenomena that affect the propagation when represented by rays. The model used is based upon classical geometric optics to work with the ray tracing engine. The scene model is comprised of a three dimensional environment in order to define the physical rotor blade surface. The receiver and transmitter will also be defined in the same environment with simulated characteristics to match real life conditions. The ray tracing engine uses recursive execution described by Turner Whitted (Whitted, 1979), replacing cameras and lights for receivers and transmitters respectively.

\section{Radio Propagation Model}
%define the math behind the physical laws that are simulated within the simulation
The propagation model defined in this project will focus on the following physical phenomena, free-space path loss, reflection, and the doppler effect. They are all formally established models described in full in (place refs here). Although several effects are described within the model several were left out to simplify the computation due to the overhead of implementation in three dimensional space and because of the assumptions made in the scene model.

@@

%formula description
where $\theta_i$ is the angle of incidence with the object's surface normal, $\varepsilon$ is the dielectric constant for that surface, and $\theta_t$ is the angle of refraction as described by Snell?s law \ref{eqn:snell}.

@@

In the following simulations performed, $\varepsilon$ will have a value of (-), which is an approximate value for (-). The final reflected power is

@@

where $P_i$ is the incident power before reflection losses.

\subsubsection{Diffuse Reflection}
A perfect diffuse reflection, also known as Lambertian reflection, will reflect rays in all directions in a uniform hemisphere (refs?). The hemisphere is located at the point of the ray hit, producing rays away from the surface. To approximate a particular type of surface the rays power is attenuated by the surface's bi-directional reflectance distribution function (BRDF) (refs?). There are several ways of calculating the BRDF, but the simplest is the Phong BRDF which will be used in this ray tracing implementation(refs?). This method for the BRDF is regularly used in the rendering of computer graphics to simulate the reflection of light off surfaces. Since is was not designed for longer wavelengths used in RF communications an assumption is made that small scale features of the surface, relevant to light waves,  produce a reflectance distribution similar for larger wavelengths. The BRDF parameters were then selected to be frequency independent to form a plausible reflectance distribution for the physical effects.

@@

\subsection{Rotor Blade}
%go over the shape of the blade
The rotor blade is the most complex object within the scene due to its airfoil shape. The airfoil shape is designed to produce lift for the helicopter, and designing one is outside the scope of this project. Fortunately there are databases that provide real helicopter airfoil designs in x,y coordinates. Theses designs form a 2D cross-section of the airfoil shape that will be extruded into the length of the rotor blade.

%figure of 2D airfoil
\begin{figure}
	\begin{center}
		\includegraphics[width=15cm]{images/radio_propagation/2d_airfoil.png}
		\caption{Airfoil}
		\label{fig:airfoil}
	\end{center}
\end{figure}

%go over how it is formed and the tessellation code
The 2D airfoil sections will form the ribs of the rotor blade but the surface of the blade will be made up of primitive surfaces to approximate the curved airfoil shape. The primitive shapes are triangles that span between the ribs and go around to cover the surface. The surface normals are then adjusted to match the correct directionality.

%figure of tessilation
\begin{figure}
	\begin{center}
		\includegraphics[width=15cm]{images/radio_propagation/blade_surface_tesselation.eps}
		\caption{Tessilation}
		\label{fig:tessilation}
	\end{center}
\end{figure}

@@

\subsection{Transmitter}
%point source transmitter casting rays in specific direction
The transmitter is modeled as a point source, this means that all the radio waves are projected from this point into the scene. It is assumed that the transmitter is located on the ground within the x,y plane in relation ship to the rotor blades. The transmitter is configured with a center frequency and is assumed to produce a tone at that frequency. Therefor signal produced has no inherent modulation applied to it.
%direction creation
Being that the Transmitter is emitting radio waves into a 3D space it is assumed to be transmitting in an omnidirectional fashion, but because the rotor and the receiver are the only two other objects within the scene radio waves will only be propagated in their direction. This limitation allows for a higher concentration of rays to be casted, resulting in a higher resolution picture and eliminates computational overhead for the tracing engine.

@@

\section{Ray Tracing}
%describe the core tracing engine and describe the flow of the program
The ray tracing simulation is performed in the Monte Carlo fashion building in the scene and propagation models. Raytracing is a well established method for creating realistic images in (ray tracing papers). The simulation is performed over successive iterations, each iteration is called a frame. Each frame will be updated to represent the change in the position of the rotor blade.

\subsection{Computation}
%flow of the program
The ray tracing engine first sets up the environment creating the surfaces and placing the objects in the scene. Then the number of frames that will be computed. The number of frames are determined by the number of rotations that the rotor makes within the simulation, the speed of the rotor in RPM and the sampling frequency of the receiver. This is described by

@@

The casting technique reduces the computation overhead by limiting the space. It also creates higher resolution result by focusing the limited number of rays in the direction of the other objects.

@@

When a ray is casted into the scene it has an origin point, a direction vector, a frequency and an initial power level. The ray then travels through the scene till it encounters another object. If there is a collision with the receiver it is recorded. If the ray collides with a rotor blade a new ray is calculated based on the normal to the geometric surface. The new ray's power is updated as well as an updated frequency since the blade is moving. The power level is adjusted based on the BRDF of the specific material. Due to only the rotor and receiver in the scene, besides the transmitter, rays that are found to hit the rotor the reflection ray is calculated. Then the angle between the reflection and a ray between the point of reflection and the receiver is used to create a dedicated ray in that direction with a decayed power and adjusted frequency.

%psudo code

%uml diagrams

>>>>

\chapter{Radio Propagation Simulation} \label{ch:radio_propagation}

\section{Introduction}
%introduce the section and what will be discussed here
The simulation software is broken down into three main components, the radio wave propagation model, the scene model, and the ray tracing engine. The wave propagation model is a collection of equations that describe the physical phenomena that affect the propagation when represented by rays. The model used is based upon classical geometric optics to work with the ray tracing engine. The scene model is comprised of a three dimensional environment in order to define the physical rotor blade surface. The receiver and transmitter will also be defined in the same environment with simulated characteristics to match real life conditions. The ray tracing engine uses recursive execution described in (Whitted, 1979), replacing cameras and lights for receivers and transmitters respectively in a three dimensional environment.

\section{Radio Propagation Model}
%define the math behind the physical laws that are simulated within the simulation
The propagation model defined in this project is based on the following physical phenomena, free-space path loss, reflection, and the doppler effect. They are all formally established models described in full in (Bertoni, 2000; Blaunstein, 2000; Griffiths, 1987)(FIND OTHERS). Although several effects are described within the model several were left out to simplify the computation due to the overhead of implementation in three dimensional space and because of the assumptions made about the scene model.

@@

%formula description
where $\theta_i$ is the angle of incidence with the object's surface normal, $\varepsilon$ is the dielectric constant for that surface, and $\theta_t$ is the angle of refraction as described by Snell's law \ref{eqn:snell}.

@@

The final reflected power is

@@

%reflection power loss description
where $P_i$ is the incident power before reflection losses.

%multipath fading
\subsection{Frequency-selective Fading}
The transmitted signal power experiences fading described by

\begin{equation}
	F = cos\left(2\pi \frac{d}{c} + t_0\right)
	\label{eqn:fading_coeff}
\end{equation}

where $F$ is the fading coefficient, $d$ is the total distance the particular ray has traveled, $c$ is the speed of light, and $t_0$ is the starting time of the current frame. The power can then be calculated using

\begin{equation}
	P_{faded} = PF
	\label{eqn:power_faded}
\end{equation}

where $P_{faded}$ is the adjusted power due to fading and $P$ is the signal power. The superposition of each ray's faded power becomes the received signal.

\subsubsection{Diffuse Reflection}
A perfect diffuse reflection, also known as Lambertian reflection, will reflect rays in all directions in a uniform hemisphere (J. Lambert, "Photometria Sive de Mensura et Gradibus Luminus Colorum et Umbrae", 1760). The hemisphere is located at the point of the ray hit, producing rays away from the surface. To approximate a particular type of surface the rays power is attenuated by the surface's bi-directional reflectance distribution function (BRDF) (Suffern, 2007; Pharr, 2010; Glassner, 1995)(OK). There are several ways of calculating the BRDF, but the simplest is the Phong BRDF which will be used in this ray tracing implementation(Phong, 1998)(OK). This method for the BRDF is regularly used in the rendering of computer graphics to simulate the reflection of light off surfaces. Since is was not designed for longer wavelengths used in RF communications an assumption is made that small scale features of the surface, relevant to light waves,  produce a reflectance distribution similar for larger wavelengths.

%The BRDF parameters were then selected to be frequency independent to form a plausible reflectance distribution for the physical effects.

@@

\subsection{Rotor Blade}
%go over the shape of the blade
The rotor blade is the most complex object within the scene due to its airfoil shape. The airfoil shape is designed to produce lift for the helicopter, and designing one is outside the scope of this project. Fortunately there are databases that provide real helicopter airfoil designs in x,y coordinates. Theses designs form a 2D cross-section of the airfoil shape that will be extruded into the length of the rotor blade. Figure \ref{fig:airfoil} shows the airfoil cross-section that will be used to create the rotor surface.

%figure of 2D airfoil
\begin{figure}
	\begin{center}
		\includegraphics[width=15cm]{images/radio_propagation/2d_airfoil.png}
		\caption{Airfoil Cross-Section}
		\label{fig:airfoil}
	\end{center}
\end{figure}

%go over how it is formed and the tessellation code
The 2D airfoil sections will form the ribs of the rotor blade but the surface of the blade will be made up of primitive surfaces to approximate the curved airfoil shape. The primitive shapes are triangles that span between the ribs and go around to cover the surface. This was done using a lightweight mesh generation method using the a priori knowledge of the blade shape. The designed algorithm, located in \ref{class_blade__surface_a160959f632ad73eff846aeea81aa84ae} $create\_surface()$ method, takes advantage of the blade being composed of rib sections with consistent shape. The surface normals are then adjusted to match the correct directionality. Figure \ref{fig:tessilation} shows the created surface for the rotor blade that will be used in the scene.

%figure of tessilation
\begin{figure}
	\begin{center}
		\includegraphics[width=15cm]{images/radio_propagation/blade_surface_tesselation.eps}
		\caption{Blade Surface Tessilation}
		\label{fig:tessilation}
	\end{center}
\end{figure}

@@

\subsection{Transmitter}
%point source transmitter casting rays in specific direction
The transmitter is modeled as a point source, this means that all the radio waves are projected from this point into the scene. It is assumed that the transmitter is located on the ground within the x,y plane in relation ship to the rotor blades. The transmitter is configured with a center frequency and is assumed to produce a tone at that frequency. Therefor signal produced has no inherent modulation applied to it.
%direction creation
Being that the Transmitter is emitting radio waves into a 3D space it is assumed to be transmitting in an omnidirectional fashion, but because the rotor and the receiver are the only two other objects within the scene radio waves will only be propagated in their direction. This limitation allows for a higher concentration of rays to be casted, resulting in a higher resolution picture and eliminates computational overhead for the tracing engine. Figure \ref{fig:transmitter_direction} shows the rays, in blue, casted only in the vicinity of the rotor blade and receiver.

@@

\section{Ray Tracing}
%describe the core tracing engine and describe the flow of the program
The ray tracing simulation is conducted in the Monte Carlo fashion acting on the scene and propagation models. This method of raytracing is established for creating realistic images in (Suffern, 2007; Pharr, 2010). The simulation is performed over successive iterations. Each iteration is a discrete element called a frame. Where each frame will be updated to represent the change in the position of the rotor blade.

\subsection{Computation}
%flow of the program
The ray tracing engine first sets up the environment creating the surfaces and placing the objects in the scene. Then the number of frames that will be computed. The number of frames is determined by the number of rotations that the rotor makes within the simulation, the speed of the rotor in RPM and the sampling frequency of the receiver defined before the beginning of the simulation. This is described by

@@

The casting technique reduces the computation overhead by limiting the search space. It also creates higher resolution result by focusing the limited number of rays in the direction of the other objects shown in figure \ref{fig:samp_method}.

@@

When a ray is casted into the scene it has an origin point, a direction vector, a frequency and an initial power level. The ray then travels through the scene till it encounters another object. If there is a collision with the receiver it is recorded. If the ray collides with a rotor blade a new ray is calculated based on the normal to the geometric surface. The new ray's power is updated as well as an updated frequency since the blade is moving. The power level is adjusted based on the BRDF of the specific material. Due to only the rotor and receiver in the scene, besides the transmitter, rays that are found to hit the rotor the reflection ray is calculated. Then the angle between the reflection and a ray between the point of reflection and the receiver is used to create a dedicated ray in that direction with a decayed power and adjusted frequency accordingly.

### 2016-10-11 Ray_Tracing_Software.tex
update files

\chapter{Ray Tracing Software}
something for code reuse and the git hub stuff

>>>>

\chapter{Ray Tracing Software}

The ray tracing software developed during the course of the thesis is written in C++ in an XCode environment, but a generic implementation is provided to run on other operating systems. This code is loosely based on the work of Eric Degroot's developed in (Eric's thesis) and implemented some modified base classes developed by him. It is a specific tool for analyzing the affects of rotor blade modulation on a signal tone, it does not provide generic 3D ray tracing in order to optimize the performance for this task. The software is provided without a GUI implementation and is run on different parameters by modifying the main.c file. This is due to the amount of parameters involved but could be easily extended to provide command line arguments in future iterations.

To use the software for performing RBM simulations there are several key parameters that are needed. First a file containing the points that define the 2D section of the airfoil shape, one default file is included in the software package for ease of use and as a template for other airfoil shapes.

the other key parameters that define the simulation are shown in table \ref{tab:key_values}

\begin{table}
\begin{center}
    \begin{tabular}{ | l | l | l | l | p{3cm} |}
    \hline
    Name & type & Default Value& units & Description\\ \hline
     rx\_x &  double & 0 & m & receiver position in x  \\ \hline
     rx\_y & double & 7 & m & receiver position in y \\ \hline 
     rx\_z & double & 2 & m & receiver position in z below rotor  \\ \hline
     Bandwidth & double & 2000 & Hz & Bandwidth of the receiver  \\ \hline
     rx\_fc & double & 1000000000 & Hz & center frequency of the receiver \\ \hline 
     tx\_x & double & 100 & m & transmitter position in x \\ \hline
     tx\_y & double & 100 & m & transmitter position in y \\ \hline
     tx\_fc & double & 1000000000 & Hz & transmitter center frequency \\ \hline
     tx\_power & double & 4000 & & transmitter power \\ \hline
     num\_blades & int & 2 & none & number of blades \\ \hline
     RPM & double & 250 & RPM & Blade revolutions per minuet \\ \hline
     altitude & double & 200 & m & rotor altitude \\ \hline
     blade\_length & double & 7.5 & m & length of the one rotor blade \\ \hline
     pitch & double & 0 & radians & pitch of the rotor use $-\theta_p$  \\ \hline
     num\_ribs & double & 10 & none & number of ribs that makes up one blade \\ \hline
     files & double & 10 & none & number of files to produce \\ \hline
    \end{tabular}
     \caption{Simulation parameters}
    \label{tab:key_values}
\end{center}
\end{table}

The output file also needs to be defined with a unique name that changes, this is important because the program will just write over the old file if you do not have a unique naming scheme. this could be just an increasing number or related to one of the variable sweeps.

To run the simulation on a Mac load into XCode the project file and configure the main file to reflect the type of simulation desired, then hit the play button.

On Linux machines the code must first be built using make with the main file configured before hand. Then run the Raytracer3D executable.

The output file contains more than just the signal data the first few data entries are data pertaining to the simulation parameters. it is important to separate this data from the signal data before processing the signal. The template for this data is shown in table 
\ref{tab:file_values}.

\begin{table}
\begin{center}
    \begin{tabular}{| l | l | l | l | l | l | l | l | l | l |}
    \hline
    1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\ \hline
      rx\_x & rx\_y & rx\_z & Bandwidth & rx\_fc & tx\_x & tx\_y & tx\_fc & tx\_power & num\_ribs \\ \hline
    \end{tabular}
    \begin{tabular}{| p{1cm} | p{1cm} | l | l | l | l | p{1.5cm} | p{1.5cm} |}
    \hline
    11 & 12 & 13 & 14 & 15 & 16 & 17 & 18\\ \hline
     num \_blades & blade \_length & pitch & RPM & altitude & tx\_radius & elevation \_angle & azimuth \_angle \\ \hline
    \end{tabular}
     \caption{Output file header with simulation parameter order}
    \label{tab:file_values}
\end{center}
\end{table}

this is all provided in a single string of comma separated values called a .CSV file. MATLAB has a built in function to read CSV data.

%matlab csv read
\begin{lstlisting}
data = csvread('file_path/file_name')
\end{lstlisting}

after all the data is read into MATLAB the data can be split into two vectors, one pertaining to the physical simulation parameters and the other to the received signal.

### 2016-10-11 Results_new.tex
update files

\chapter{Results}

@@

\section{Transmitter Azimuth Angle Estimation}
%formulate the estimation equation and discuss ambiguity considerations
The azimuth angle estimation technique is based on the data collected by rotating the helicopter platform in a clockwise manner, during which at a set angle a doppler envelope is calculated and subsequent maximum and minimum values are collected. After a complete rotation the doppler maximum and minimum values produce a sinusoidal pattern as shown in figure \ref{fig:tx_azimuth_angle_200} without the assumption of knowing the starting azimuth angle. With this data we need to find an unambiguous angle that will set the coordinate space and provide an estimate of current azimuth angle. To do this the absolute value of the minimum envelope data is superimposed onto the maximum envelope data as shown in figure \ref{fig:azimuth_estimation_max_vs_absMin}.

@@

\section{Transmitter Elevation Angle Estimation}
%formulate the estimation equation and discuss ambiguity considerations
Using the information gathered from the azimuth estimation procedure we can compute the elevation angle of the transmitter. In the process of gathering azimuth information the maximum and minimum envelopes follow a sinusoidal pattern that contains a maximum Doppler information at approximately 180\textdegree \space and 0\textdegree \space azimuth angles. Using that maximum value along with the correct Doppler and the assumption that the maximum will occur near the tip of the rotor blade we can approximate the respected reflection radii off the blade as the azimuth angle changes. This is done through 
%reflection radius aproximation equation
\begin{equation}
	R_{Estimate} = BladeLength \left(\frac{f_{DopplerCorrect}(\theta_{Az})}{Max(f_{DopplerCorrect})}\right)
	\label{eqn:r_estimate}
\end{equation}

@@

@@

This error is caused by a limitation on the estimation is that the doppler shift is limited by the maximum amount doppler that the blade rotation can produce shown in equation (-) by replacing $v_r$ with $v_rMax$ which is calculated by

@@

>>>>

\chapter{Results} \label{ch:results}

@@

\section{Transmitter Azimuth Angle Estimation} \label{sec:transmitter_azimuth_angle_estimation}
%formulate the estimation equation and discuss ambiguity considerations
The azimuth angle estimation technique is based on the data collected by rotating the helicopter platform in a clockwise manner, during which at a set angle a doppler envelope is calculated and subsequent maximum and minimum values are collected. After a complete rotation the doppler maximum and minimum values produce a sinusoidal pattern as shown in figure \ref{fig:tx_azimuth_angle_200} without the assumption of knowing the starting azimuth angle. With this data we need to find an unambiguous angle that will set the coordinate space and provide an estimate of current azimuth angle. To do this the absolute value of the minimum envelope data is superimposed onto the maximum envelope data as shown in figure \ref{fig:azimuth_estimation_max_vs_absMin}.

@@

\section{Transmitter Elevation Angle Estimation} \label{sec:transmitter_elevation_angle_estimation}
%formulate the estimation equation and discuss ambiguity considerations
Using the information gathered from the azimuth estimation procedure we can compute the elevation angle of the transmitter. In the process of gathering azimuth information the maximum and minimum envelopes follow a sinusoidal pattern that contains a maximum Doppler information at approximately 180\textdegree \space and 0\textdegree \space azimuth angles. Using that maximum value along with the correct Doppler and the assumption that the maximum will occur near the tip of the rotor blade we can approximate the respected reflection radii off the blade as the azimuth angle changes. This is done through 
%reflection radius aproximation equation
\begin{equation}
	R_{Estimate} = BladeLength \left(\frac{f_{DopplerCorrect}(\theta_{Az})}{Max(f_{DopplerCorrect})}\right)
	\label{eqn:r_estimate}
\end{equation}

@@

where $f_{DopplerMax}$ is defined in equation \ref{eqn:fd_max}.

@@

This error is caused by a limitation on the estimation is that the doppler shift is limited by the maximum amount doppler that the blade rotation can produce shown in equation \ref{eqn:v_radial} by replacing $v_r$ with $v_rMax$ which is calculated by

@@

%------------------------------
%pitch considerations
%------------------------------

If we now consider a pitched blade the above method for estimating Azimuth and elevation becomes inaccurate due to the increased doppler between 90\textdegree and 270\textdegree and the decreased doppler on the other side. To correct this error we need to correct both the difference calculation and the correct doppler data accordingly. The uncorrected data is shown in figure \ref{fig:diff_pitch_noCorrection} and \ref{fig:dop_pitch_noCorrection} with percent azimuth error of 20\% and percent elevation error of 35\%.

\begin{figure}
	\begin{center}
		\includegraphics[width=10cm]{images/results/difference_pitch_noCorrection.eps}
		\caption{Uncorrected Envelope Difference vs. Transmitter Azimuth and Elevation angle of 45\textdegree with Blade pitch of 7.5\textdegree}
		\label{fig:diff_pitch_noCorrection}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=10cm]{images/results/correct_dop_pitch_noCorrection.eps}
		\caption{Uncorrected Doppler vs. Transmitter Azimuth and Elevation angle of 45\textdegree with Blade pitch of 7.5\textdegree}
		\label{fig:dop_pitch_noCorrection}
	\end{center}
\end{figure}

Comparing the envelope difference of no pitch and pitched blade, in figures \ref{fig:azimuth_estimation_difference} and \ref{fig:diff_pitch_noCorrection} respectively, we can see that the difference calculation of the pitch is a shifted up version of the one with no pitch. Therefore we can first correct this by re-centering the pitch difference to be located around zero.

\begin{equation}
	A_{shift} = \frac{Max(EnvelopeDifference) - |Min(EnvelopeDifference)|}{2}
	\label{eqn:pitch_shift}
\end{equation}

\begin{equation}
	EnvelopeDifference = EnvelopeDifference - A_{shift}
	\label{eqn:diff_shift}
\end{equation}

Then with this knowledge $f_{DopplerCorrect}$ is recalculated and then is adjusted to the correct frequencies with

%fix correct doppler for pitch
\begin{equation}
	f_{DopplerCorrect} = f_{DopplerCorrect} + A_{shift}cos\left(\theta_{Az}\right)
	\label{eqn:fd_max}
\end{equation}

Since $\theta_{Az}$ is unknown during the calculation of this equation the correct starting phase can be determined by matching it to the $EnvelopeDifference$ with the a priori information that the cosign should be one period in length. These corrections result in the percent azimuth error of 10\% and 6.5\% elevation angle error. The error in the azimuth is quite large compared to the results when pitch is 0\textdegree but the sample is only comprised of 41 points which turns out to be 9\textdegree per sample in which 10\% error would be off by one sample. Increasing the samples per revolution would most likely decrease this error. The elevation angle percent error is within a useable range but could be decreased with further tuning to the preliminary pitch adjustment algorithm.
